---
title: "STAT-627 Space: KNN 2"
author: "Max Calzada, Josiah Gottfried, Sanghyeob Ko, Domingo Salerno"
date: '2022-12-03'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(class)
library(data.table)
library(dplyr)
library(pls)
library(tidyr)
library(tidyverse)
```

# Introduction

For our project, our central question was "Which statistical model would most optimally classify whether a celestial body is a star, quasar, or galaxy."

To answer this question we built the following models:

KNN
LDA/QDA
Logistic Regression
PCA
Trees/Random Forests

## Data

The data consists of 100,000 observations of space taken by the SDSS (Sloan Digital Sky Survey). Every observation is 
described by 17 feature columns and 1 class column which identifies it to be either a star, galaxy or quasar.

<https://www.kaggle.com/datasets/fedesoriano/stellar-classification-dataset-sdss17>

# Models

## KNN

```{r}
setwd("~/Desktop/STAT-627/STAT627_Space_Project")

Space_DF <- read.csv("star_classification.csv")
Space_DF <- Space_DF[, c("class", "u", "r", "g", "i", "z", "redshift")]
Space_DF$class_f <- as.factor(Space_DF$class) # Class or Factor? Why?
```

```{r}
set.seed(1)
n.S <- nrow(Space_DF) 
Z.S <- sample(n.S, n.S*0.9) # We’ll split data at random 
```

```{r}
S.Train <- Space_DF[Z.S,] # training data
S.Test <- Space_DF[-Z.S,] # testing data

X.S.Train <- S.Train[, c("u", "r", "g", "i", "z", "redshift")]
X.S.Test <- S.Test[, c("u", "r", "g", "i", "z", "redshift")]
Y.S.Train <- Space_DF$class_f[Z.S]
Y.S.Test <- Space_DF$class_f[-Z.S]
```

```{r}
knn.S = knn(X.S.Train, X.S.Test, Y.S.Train, 3) # K = 3?
table(Y.S.Test, knn.S)
```

```{r}
mean(Y.S.Test == knn.S )
```

```{r}
class.rate = rep(0,20) # Create a vector of length 20 and fill it with classification rates,
# which number instead of 20?
```

```{r}
for (K in 1:20) {
  knn.S = knn(X.S.Train, X.S.Test, Y.S.Train, K)
  class.rate[K] = mean(Y.S.Test == knn.S)
}

class.rate
```

```{r}
which.max(class.rate)
max(class.rate)
```

```{r}
# K20 <- as.vector(1:20)
# plot(K,class.rate)
plot(class.rate)
```

## LDA/QDA
1. Data Analysis 1 (LDA)
1-1. Linear Discriminant Analysis (LDA)
```{r}
# library(MASS)
LDA <- lda(class ~ ., data = Space_DF) # The main command for LDA
LDA
# Prior probabilities of groups: These are sample proportions of the 3 groups, from our data
# Group means: Multivariate group means are computed within each group
# Coefficients of linear discriminants: For our information only: these functions LD1-LD2
# LD1, LD2 are different from our linear discriminant functions. 
# These printed coefficients determine the Fisher's linear discriminants LD1, LD2. The first one is a linear function that achieves the maximal separation of our four groups. LD2 is a linear function, orthogonal to LD1, that achieves the maximal separation among all linear functions orthogonal to LD1, etc. These functions are linear combinations of our linear discriminant functions. Their derivation is based on Linear Algebra. Here, LD1 captures 87.93% of differences between the groups, LD2 adds 12.07% to that.
```

1-2. LDA Cross-validation (CV: LOOCV)
```{r}
# Cross-validation
# Option CV=TRUE is used for “leave one out” cross-validation; for each sampling unit, it gives its class assignment without the current observation. This is a method of estimating the testing classifications rate instead of the training rate.
LDA.fit <- lda(class ~., CV = TRUE, data = Space_DF)
table(Space_DF$class, LDA.fit$class) # The main diagonal shows correctly classified counts.
mean(Space_DF$class == LDA.fit$class) # 84.62%: Correct classification rate = proportion of correctly classified counts.
```

1-3. LDA Cross-validation (CV: Cross-validation, training:testing = 9:1)
```{r}
LDA.training <- lda(class ~ ., data = S.Train) # Only used the training set
LDA.pred <- predict(LDA.training, newdata = S.Test) # Prediction by using the testing set

table(S.Test$class, LDA.pred$class) # The main diagonal shows correctly classified counts.
mean(S.Test$class == LDA.pred$class) # 84.67%: Correct classification rate = proportion of correctly classified counts.
```


1-4. LDA Cross-validation (CV: k-fold 10)

```{r}
fold <- sample(rep(1:10, 10000), n)
pred_rate <- NULL

for(i in 1:10){
  train_kf <- Space_DF[fold != i,]
  test_kf <- Space_DF[fold == i,]
  lda_kf <- lda(class ~ ., data = train_kf )
  lda_kf_pred <- predict(lda_kf, test_kf)
  pred_rate[i] <- mean(test_kf$class == lda_kf_pred$class)
}

mean(pred_rate) # 84.63% for K-fold 10 validation of LDA
```


2. Data Analysis 2 QDA
2-1. Quadratic Discriminant Analysis (QDA)
```{r}
QDA <- qda(class ~ ., data = Space_DF) # The main command for QDA
QDA
```


2-2. QDA Cross-validation (CV: LOOCV)
```{r}
# Quadratic Discriminant Analysis
QDA.fit <- qda(class ~ ., CV = TRUE, data = Space_DF)
table(Space_DF$class, QDA.fit$class) # The main diagonal shows correctly classified counts.
mean(Space_DF$class == QDA.fit$class) # NA
(55860+17277+21410)/(55860+3104+478+1677+17277+2+63+120+21410) # 94.55%: Correct classification rate = proportion of correctly classified counts.
```


2-3. QDA Cross-validation (CV: Cross-validation, training:testing = 9:1)
```{r}
QDA.training <- qda(class ~ ., data = S.Train) # Only used the training set
QDA.pred <- predict(QDA.training, newdata = S.Test) # Prediction by using the testing set

table(S.Test$class, QDA.pred$class) # The main diagonal shows correctly classified counts.
mean(S.Test$class == QDA.pred$class) # 94.54%: Correct classification rate = proportion of correctly classified counts.
```



2-4. QDA Cross-validation (CV: k-fold 10)

```{r}
fold <- sample(rep(1:10, 10000), n)
pred_rate <- NULL

for(i in 1:10){
  train_kf <- Space_DF[fold != i,]
  test_kf <- Space_DF[fold == i,]
  qda_kf <- qda(class ~ ., data = train_kf )
  qda_kf_pred <- predict(qda_kf, test_kf)
  pred_rate[i] <- mean(test_kf$class == qda_kf_pred$class)
}
mean(pred_rate) # 94.56% for K-fold 10 validation of QDA
```



4. LDA & QDA with PC (PC1, PC2)

```{r}
X.pca <-  model.matrix(class ~ u + r + g + i + z + redshift, data = Space_DF)[,-1]
pc <-  prcomp(X.pca, scale = TRUE)
summary(pc)
screeplot(pc)
pc_df <- data.frame(Space_DF$class, pc$x)
pc_df <- rename(pc_df, 'class' = 'Space_DF.class')
pc_df

# Training set / Testing set of PC data
n <- length(pc_df$class)
n
z <- sample(n, n*0.9)
S.Train.PC <- pc_df[z,]
S.Test.PC <- pc_df[-z,]
```

4-1. LDA with PC
```{r}
LDA.PC <- lda(class ~ PC1 + PC2 + PC3, data = pc_df)
LDA.PC 
# LD1 captures 91.08% of differences between the groups, LD2 adds 8.92% to that. The result is better than before. Originally, LD1 captured 87.93% of differences between the groups, LD2 adds 12.07% to that.
```


4-2. LDA with PC Cross-validation (CV: LOOCV)
```{r}
LDA.fit.PC <- lda(class ~ PC1 + PC2 + PC3, CV = TRUE, data = pc_df)
table(pc_df$class, LDA.fit.PC$class) 
mean(pc_df$class == LDA.fit.PC$class) # 82.71%. It is a little bit lower than original  84.62%, correct classification rate.
```

4-3. LDA with PC Cross-validation (CV: Cross-validation, training:testing = 9:1)
```{r}
LDA.training.PC <- lda(class ~ PC1 + PC2 + PC3, data = S.Train.PC)
LDA.pred.PC <- predict(LDA.training.PC, newdata = S.Test.PC)

table(S.Test.PC$class, LDA.pred.PC$class)
mean(S.Test.PC$class == LDA.pred.PC$class) # 82.17%. It is a little bit lower than original 84.67%, correct classification rate.
```


4-4. LDA with PC Cross-validation (CV: k-fold 10)

```{r}
fold <- sample(rep(1:10, 10000), n)
pred_rate <- NULL

for(i in 1:10){
  train_kf <- pc_df[fold != i,]
  test_kf <- pc_df[fold == i,]
  lda_kf <- lda(class ~ PC1 + PC2 + PC3, data = train_kf )
  lda_kf_pred <- predict(lda_kf, test_kf)
  pred_rate[i] <- mean(test_kf$class == lda_kf_pred$class)
}

mean(pred_rate) # 82.72%, It is a little bit lower than original 84.63% for K-fold 10 validation of LDA
```



4-5. QDA with PC
```{r}
QDA.PC <- qda(class ~ PC1 + PC2 + PC3, data = pc_df) 
QDA.PC
```


4-6. QDA with PC Cross-validation (CV: LOOCV)
```{r}
QDA.fit.PC <- qda(class ~ PC1 + PC2 + PC3, CV = TRUE, data = pc_df)
table(pc_df$class, QDA.fit.PC$class)
mean(pc_df$class == QDA.fit.PC$class) # 94.55%. It is almost the same as the original 94.55%
```


4-7. QDA with PC Cross-validation (CV: Cross-validation, training:testing = 9:1)
```{r}
QDA.training.PC <- qda(class ~ PC1 + PC2 + PC3, data = S.Train.PC)
QDA.pred.PC <- predict(QDA.training.PC, newdata = S.Test.PC)

table(S.Test.PC$class, QDA.pred.PC$class)
mean(S.Test.PC$class == QDA.pred.PC$class) # 94.49%, it is almost the same as the original 94.54%.
```



4-8. QDA with PC Cross-validation (CV: k-fold 10)

```{r}
fold <- sample(rep(1:10, 10000), n)
pred_rate <- NULL

for(i in 1:10){
  train_kf <- pc_df[fold != i,]
  test_kf <- pc_df[fold == i,]
  qda_kf <- qda(class ~ PC1 + PC2 + PC3, data = train_kf )
  qda_kf_pred <- predict(qda_kf, test_kf)
  pred_rate[i] <- mean(test_kf$class == qda_kf_pred$class)
}
mean(pred_rate) # 94.55%. It is almost the same as the original 94.56%
```



## Logistic Regression

## PCA

```{r}
setwd("~/Desktop/STAT-627/STAT627_Space_Project")

Space_DF <- read.csv("star_classification.csv")
Space_DF <- Space_DF[, c("class", "u", "r", "g", "i", "z", "redshift")]

Space_DF <- Space_DF %>%
  mutate(class = as.factor(class)) %>%
  subset(u > min(u))
# Space_DF$class_f <- as.factor(Space_DF$class)
```

```{r}
X.pca = model.matrix(class ~ u + r + g + i + z + redshift, data = Space_DF)[,-1]
pc = prcomp(X.pca, scale = TRUE)
summary(pc)
screeplot(pc)
```

## Trees/Random Forests

# Results

The model with the best prediction rate was the random forest model at 97.76% accuracy.
